---
title: "Sandhills Grassland Birds groups as strata  🐦"
author: Eric Rexstad *University of St Andrews*
date: "`r Sys.Date()`"
date-format: "DD MMMM YYYY"
number-sections: true
bibliography: rachel.bib
csl: ../apa.csl
---

:::{.callout-note collapse=false appearance='default' icon=true}
## Second analysis of grassland birds
- Alternative definition of strata
  - Previously 33 strata were defined
- This analysis uses "group" as the stratification criterion
  - Resulting in 19 strata, with a resulting increase in number of detections underpinning the 19 estimates of density
:::


# Becoming acquainted with data

I will also make the adjustment with `Group` becoming `Region.Label` and `GroupArea` becoming `Area`
```{r}
alldata <- read.csv(file="data/Rusten_GrasslandBird_dist24_NEW.csv")
colnames(alldata)[colnames(alldata) == 'Region.Label'] <- 'formerRegion.Label'
colnames(alldata)[colnames(alldata) == 'Area'] <- 'formerArea'
colnames(alldata)[colnames(alldata) == 'Group'] <- 'Region.Label'
colnames(alldata)[colnames(alldata) == 'GroupArea'] <- 'Area'
(namstrat <- unique(alldata$Region.Label))
(namspec <- unique(alldata$Species))
```

Goodness, there are `r length(namspec)` species recorded during the 2024 field season. Let's focus attention upon three species of interest: GRSP, WEME, DICK:

```{r}
threespec <- subset(alldata, subset = Species %in% c("GRSP", "WEME", "DICK"))
knitr::kable(table(threespec$Region.Label, threespec$Species),
             caption="This table identical to second table in report of first analysis.")
```

## Transects within strata

The revised data has placed transect labels in the field `Sample.Label` where the labels should be. This is a departure from the previous data set when the transect labels resided in a (now removed) field `GridPointID`. That makes life easier, as there is no field renaming to perform.

Problems can arise when subsetting the data, with concomitant loss of information about the survey design. The inclusion of a record for each point transect (using NA for species and distance) maintains the integrity of the design within the flatfile. The next code chunk creates a `sample table` that encapsulates the design information such that the data can be subset (by species for example), while having the survey design information available during the analysis.

```{r}
GroupArea <- mat.or.vec(length(namstrat),1)
j <- 0
for (i in namstrat) {
  j <- j+1
  GroupArea[j] <- unique(alldata[alldata$Region.Label==i, "Area"])
}
regionTable <- data.frame(Region.Label=namstrat, Area=GroupArea)

numtran <- mat.or.vec(length(namstrat),1)
namtran <- list()
j <- 0
for (i in namstrat) {
  j <- j+1
  numtran[j] <- length(unique(alldata[alldata$Region.Label==i, "Sample.Label"]))
  namtran[[j]] <- unique(alldata[alldata$Region.Label==i, "Sample.Label"])
#  print(paste(i, numtran[j], sep=" "))
}
repeatGroups <- c(rep(namstrat[1], numtran[1]),
                  rep(namstrat[2], numtran[2]),
                  rep(namstrat[3], numtran[3]),
                  rep(namstrat[4], numtran[4]),
                  rep(namstrat[5], numtran[5]),
                  rep(namstrat[6], numtran[6]),
                  rep(namstrat[7], numtran[7]),
                  rep(namstrat[8], numtran[8]),
                  rep(namstrat[9], numtran[9]),
                  rep(namstrat[10], numtran[10]),
                  rep(namstrat[11], numtran[11]),
                  rep(namstrat[12], numtran[12]),
                  rep(namstrat[13], numtran[13]),
                  rep(namstrat[14], numtran[14]),
                  rep(namstrat[15], numtran[15]),
                  rep(namstrat[16], numtran[16]),
                  rep(namstrat[17], numtran[17]),
                  rep(namstrat[18], numtran[18]),
                  rep(namstrat[19], numtran[19]))
pointStations <- c(namtran, recursive=TRUE)
sampleTable <- data.frame(Sample.Label=pointStations,
                          Region.Label=repeatGroups,
                          Effort=1)
```

:::{.callout-caution collapse=false appearance='default' icon=true}
## Distances recorded as -1
Don't know why they exist, but they mess with detection function fitting, I'm throwing them out
:::

```{r}
alldata <- subset(alldata, subset = distance> -1)
```


With data wrangling complete, we can begin our model fitting. I do not repeat the histogram of detection distances by common species, as they can be seen in the previous report.

# Simple analysis for estimates at the `Group` level

Standard practice will be to remove 10% most distant detections.

```{r}
#| label: loadpkg
#| message: false
library(Distance)
cu <- convert_units("meter", NULL, "hectare")
```

To use on several species, make a small function to fit 3 candidate models and apply to *Western meadowlark*. By eye, I've chosen an absolute truncation distance of 170m.

```{r}
#| label: threecand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, WEME, pooled over groups"
three <- function(dframe=alldata, spec, trunpct="10%", 
                  rt=regionTable, st=sampleTable) {
  thisspec <- subset(dframe, Species==spec)
  this.hn <- ds(data=thisspec, transect = "point",
                region_table = rt, sample_table = st,
                key="hn", truncation = trunpct,  convert_units = cu, 
                adjustment = "herm")
  this.uncos <- ds(data=thisspec, transect = "point",
                region_table = rt, sample_table = st,
                   key="unif", truncation = trunpct, convert_units = cu)
  this.hr <- ds(data=thisspec, transect = "point",
                region_table = rt, sample_table = st,
                key="hr", truncation = trunpct,  convert_units = cu,
                adjustment = "poly")
  criticism <- summarize_ds_models(this.hn, this.uncos, this.hr)
  print(knitr::kable(criticism[,2:7], row.names = FALSE, digits=3,
               caption=paste("Model criticism for", spec)))
  mytitle <- paste(spec, criticism$'Key function'[1], " pooled over Groups")
  if(grepl("uncos", criticism$Model[1])) plot(this.uncos, pdf=TRUE, nc=40, main=mytitle)
  if(grepl("hn", criticism$Model[1])) plot(this.hn, pdf=TRUE, nc=40,main=mytitle)
  if(grepl("hr", criticism$Model[1])) plot(this.hr, pdf=TRUE, nc=40,main=mytitle)
  return(list(hn=this.hn, unicos=this.uncos, hr=this.hr))
}
wemeres <- three(alldata, "WEME", 170, rt=regionTable, st=sampleTable)
```


```{r}
#| label: grspcand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, GRSP, pooled over groups"
grspres <- three(alldata, "GRSP", "10%", rt=regionTable, st=sampleTable)
```

```{r}
#| label: dickcand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, DICK, pooled over groups"
dickres <- three(alldata, "DICK", "10%", rt=regionTable, st=sampleTable)
```

# Constant detectability across strata?

Clearly there are insufficient detections within strata to fit stratum-specific detection functions. However, as shown in @rexstad2023, bias will result from using a pooled detection function to estimate density at the stratum level **if** detectability differs between strata.  

The defensible way to estimate stratum-specific detection probabilities with small stratum-specific detections is to use stratum as a covariate. This may not work for the large number of strata we have here. The hazard rate model was preferred for two of the three species and it was second-preferred for the GRSP, so attempt covariate modelling using hazard rate as key function.

```{r}
#| message: false
#| label: regioncovariate
weme.hrstrat <- ds(data=alldata[alldata$Species=="WEME",], transect="point", 
                   key="hr", truncation=170,
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
grsp.hrstrat <- ds(data=alldata[alldata$Species=="GRSP",], transect="point", 
                   key="hr", truncation="10%",
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
dick.hrstrat <- ds(data=alldata[alldata$Species=="DICK",], transect="point", 
                   key="hr", truncation="10%",
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
```

:::{.callout-warning collapse=false appearance='default' icon=true}
## Convergence warning)
There are convergence warnings for all three species when using stratum as a covariate
:::

Contrast models with and without Region.Label covariate for each species to provide an evidence-based assessment of constant detectability across strata.

```{r}
#| label: aic
AIC(wemeres[["hr"]], weme.hrstrat)
AIC(dickres[["hr"]], dick.hrstrat)
AIC(grspres[["hr"]], grsp.hrstrat)
```

These AIC comparisons indicate that there is some evidence to suggest detectability differs between Groups for each of the three species.

We could also use BIC to assess the effect of Group on detection probability

```{r}
#| label: bic
altmetric <- function(modelobj) {
  nval <- length(modelobj$ddf$fitted)
  AICcall <- AIC(modelobj)
  k <- AICcall$df
  AICval <- AICcall$AIC
  AICc <- AICval + (2*k*(k+1)) / (nval-k-1)
  BIC <- AIC(modelobj, k = log(nval))$AIC
  return(BIC)
}
wconst <- altmetric(wemeres[["hr"]])
wregion <- altmetric(weme.hrstrat)
dconst <- altmetric(dickres[["hr"]])
dregion <- altmetric(dick.hrstrat)
gconst <- altmetric(grspres[["hr"]])
gregion <- altmetric(grsp.hrstrat)
bictab <- data.frame(constant=c(wconst, dconst, gconst),
                     Groupcov=c(wregion, dregion, gregion))
row.names(bictab) <- c("WEME", "DICK", "GRSP")
knitr::kable(bictab, caption="BIC comparison for Region.Label effect on detectability for three species. BIC prefers a model of constant detectability across Region.Labels.")
```

The BIC findings contradict the AIC findings of a Region.Label effect upon detection probability. BIC places a heavier penalty as number of detections increases. In reality, the differences in the shapes of the region-specific detection functions are not substantial; meaning that if detection probabilites differ between regions, that difference is small.

# Density estimates by Group, assuming constant detectability across strata

```{r}
knitr::kable(wemeres[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of WEME detections")
knitr::kable(dickres[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of DICK detections")
knitr::kable(grspres[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of GRSP detections")
```

```{r}
knitr::kable(wemeres$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for WEME")
knitr::kable(dickres$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for DICK")
knitr::kable(grspres$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for GRSP")
```

# Other species, besides the three thusfar assessed

```{r}
detects <- sort(table(alldata$Species))
knitr::kable(rev(detects[detects>200]))
```

I'll carry on through BHCO, FISP, MODO and LASP; remaining species have average detections fewer than 10 per Group.

```{r}
#| label: bhcocand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, BHCO, pooled over groups"
bhcores <- three(alldata, "BHCO", "10%", rt=regionTable, st=sampleTable)
```

```{r}
#| label: fispcand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, FISP, pooled over groups"
fispres <- three(alldata, "FISP", "10%", rt=regionTable, st=sampleTable)
```

```{r}
#| label: modocand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, MODO, pooled over groups"
modores <- three(alldata, "MODO", "10%", rt=regionTable, st=sampleTable)
```

```{r}
#| label: laspcand
#| message: false
#| warning: false
#| fig-cap: "Fitted detection function, LASP, pooled over groups"
laspres <- three(alldata, "LASP", "10%", rt=regionTable, st=sampleTable)
```

And fitting Region.Label covariate models

```{r}
#| message: false
#| label: regioncovariate4
bhco.hrstrat <- ds(data=alldata[alldata$Species=="BHCO",], transect="point", 
                   key="hr", truncation=170,
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
fisp.hrstrat <- ds(data=alldata[alldata$Species=="FISP",], transect="point", 
                   key="hr", truncation="10%",
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
modo.hrstrat <- ds(data=alldata[alldata$Species=="MODO",], transect="point", 
                   key="hr", truncation="10%",
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
lasp.hrstrat <- ds(data=alldata[alldata$Species=="LASP",], transect="point", 
                   key="hr", truncation="10%",
                   region_table = regionTable, sample_table = sampleTable,
                   convert_units = cu, formula=~Region.Label)
```


Finally BIC model selection for hazard rate models with and without a Region.Label covariate for the final four species:

```{r}
#| message: false
#| label: regioncov4
bconst <- altmetric(bhcores[["hr"]])
bregion <- altmetric(bhco.hrstrat)
fconst <- altmetric(fispres[["hr"]])
fregion <- altmetric(fisp.hrstrat)
mconst <- altmetric(modores[["hr"]])
mregion <- altmetric(modo.hrstrat)
lconst <- altmetric(laspres[["hr"]])
lregion <- altmetric(lasp.hrstrat)
bictab <- data.frame(constant=c(bconst, fconst, mconst, lconst),
                     Groupcov=c(bregion, fregion, mregion, lregion))
row.names(bictab) <- c("BHCO", "FISP", "MODO", "LASP")
knitr::kable(bictab, caption="BIC comparison for Region.Label effect on detectability for remaining four species. BIC prefers a model of constant detectability across Region.Labels.")
```

```{r}
knitr::kable(bhcores[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of BHCO detections")
knitr::kable(fispres[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of FISP detections")
knitr::kable(modores[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of MODO detections")
knitr::kable(laspres[["hr"]][["dht"]][["individuals"]][["summary"]], digits=3,
             caption="Summary of LASP detections")
```

```{r}
knitr::kable(bhcores$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for BHCO")
knitr::kable(fispres$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for FISP")
knitr::kable(modores$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for MODO")
knitr::kable(laspres$hr$dht$individuals$D, digits=3,
             caption="Density estimates by Group for LASP")
```
